---
title: "An overview of time-to-event analysis for dental researchers"
author: "Tabitha K. Peter"
date: \today
output: pdf_document
bibliography: references/cmr.bib
header-includes: 
 - \usepackage{bbm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	collapse = TRUE
)
```

Welcome to the tutorial "An overview of time-to-event analysis for dental researchers." The objective of this tutorial is to explain the foundational concepts of time-to-event analysis for an audience of dental researchers. 

These foundational concepts are illustrated using an extended example from the dental research literature. See [this link](https://www.sciencedirect.com/science/article/pii/S000281772100547X?casa_token=criQNRkNLBEAAAAA:nL6pVTDH07s1jyh5Gf_J1Ffsz2JDTmDakeGvRlgAB_rKSkjVynSdt5c_KDlZw7hlDphCLKoH) for the full text of this published work. 


# Set up 

First, load the data set 

```{r source_calls}
source("R/R.R")
cmr <- read_csv("data/cmr.csv")
```

Once the data is loaded, take a look to see the contents of the data set. I notice that there are a lot of options for 'clinic' -- this tells me I will need to combine some categories before analyzing the data. 

```{r first_look}
# First look at the data 
# See what changes we need to make 
dplyr::glimpse(cmr) # notice: lots of options for 'clinic', and some categories are labeled as numbers! 
table(cmr$Clinic, useNA = "always") # hmmm... need to combine some categories 

```

I make some changes to the formatting of the data, so that I am ready to analyze the data. The goal of this analysis is to describe what factors influence the length of time that a crown margin repair lasts. 

```{r data_prep}
# Make some formatting changes
cmr <- cmr %>%
  # collapse categories for clinic (we will need this for analysis)
  mutate(Clinic = if_else(Clinic %in% c("ADMS", "CCST", "CODHDG", "FEECLS", 
                                        "GMU", "HDGEN", "HDPROS", "LTDCAR"),
                          "Other",
                          Clinic),
         # tell the computer that 'No_surface' numbers represent *categories* 
         No_Surfaces = as.character(No_Surfaces),
         # tell the computer that RCT is a yes/no outcome
         RCT = case_when(
           RCT == 0 ~  "No RCT",
           RCT == 1 ~ "RCT"),
         # make the IDs have readable names 
         RecordID = paste0("Patient ", RecordID),
         # Represent the outcome (reintervention) in both numbers and words
         Event = Status, 
         Status = case_when(
           Event == 0 ~ "Censored",
           Event == 1 ~ "Event")) 

```


# Descriptive analysis 

Once I have verified that the data set is ready for analysis, I create a "Table 1" that summarizes each variable in the data set. 

```{r descriptives}

# Create Table 1 
table1 <- tableby(~ .,
                  data = cmr %>%
                    dplyr::select(-c(RecordID, Event, No_Cases, Tooth_Number,
                                     Date_Repair, End_Date, Failure_Date)),
                  control = tableby.control(
                    test = FALSE
                  )
                  )

# Print out Table 1 in a readable format 
summary(table1) %>% 
  kable(digits = 2,
        format = "pipe",
        caption = "Table 1: Description of Data")





```


## Timeline chart 

To illustrate the stories of specific patients, I draw a timeline chart (some people call this type of graph a "swimlane diagram"). 

```{r timeline}
# Create a timeline chart 
set.seed(52246) # this means the same patients are chosen each time I run this code
example_patients <- sample(1:nrow(cmr), size = 5)

timelines <- ggplot(cmr[example_patients, ], aes(x = RecordID, y = Time)) + 
  geom_col(position = position_dodge(), width = 0.1, alpha = 0.5) +
  geom_point(data = cmr[example_patients, ],
             aes(RecordID, Time, color = Status, shape = Status),
             size = 6) + 
  coord_flip() + 
  scale_color_viridis(discrete = TRUE) + 
  theme_bw()


(timelines)

```

## Kaplan-Meier plot 

To get an idea of the overall trajectory of the time that the crown margin repairs are lasting, a create a Kaplan-Meier (KM) plot. As time goes on, there are fewer repairs upon which to draw estimates, so our estimates become more uncertain at later years. To visualize this uncertainty, we can add confidence intervals to the plots. A risk table provides details to supplement the general pattern illustrated in the KM plot. 

```{r km_plot}

# Kaplan Meier (KM) curve 
km1 <- survfit2(Surv(time = Time, event = Event) ~ 1, data = cmr) %>% 
  ggsurvfit() +
  labs(
    x = "Years (after repair)",
    y = "Overall probability that repair is intact"
  ) 
km1

# KM curve with confidence intervals and risk table 
km2 <- survfit2(Surv(time = Time, event = Event) ~ 1, data = cmr) %>% 
  ggsurvfit() +
  labs(
    x = "Years (after repair)",
    y = "Overall probability that repair is intact"
  ) + 
  add_confidence_interval() + 
  add_risktable()

km2

```


# Analysis 

## Median survival 

**Remember:** The median survival time is **not** just the median of all the survival time values. When we are talking about median survival, we have to account for the fact that some repairs are censored -- we do not know everything about each repair in our study! The Kaplan-Meier method for calculating median survival takes censoring into account. We can do this calculation in `R` with the `survfit` function from the `survival` package. 

```{r median_survival}
# calculate median survival with the KM method 
s1 <- survfit(Surv(Time, Event) ~ 1, data = cmr) 

# create a table with the median survival time and its accompanying 95% confidence interval
s1_median <- summary(s1)$table

s1_median[c("median", "0.95LCL", "0.95UCL")] %>%
  t() %>%
 kable(digits = 3,
       col.names = c(
         "Median survival time",
         "95% CI (lower)",
         "95% CI (upper)"
       ))
```

## Nth year survival 

We are also often interested to estiamte the survival probability of a repair making it __ number of years. Below, I use the same `survfit` function to estimate 1, 3, and 5 year survival with the KM method. 

**NB**: to make the following tables have a readable format, I wrote my own `R` function `nth_yr_surv()` - the code for this function is the `data\R.R` file. 

```{r nth_year_survival}
# one year survival -------------------------------------------------
one_yr_surv <- summary(survfit(Surv(Time, Event) ~ 1, data = cmr), times = 1)

nth_year_survival(one_yr_surv) %>% kable(digits = 3)

# 1 and 3 year survival ----------------------------------------------
three_yr_surv <- summary(survfit(Surv(Time, Event) ~ 1, data = cmr), times = c(1, 3))

nth_year_survival(three_yr_surv) %>% kable(digits = 3)


# 1, 3, and 5 year survival -----------------------------------------------
five_yr_surv <- summary(survfit(Surv(Time, Event) ~ 1, data = cmr), times = c(1, 3, 5))

nth_year_survival(five_yr_surv) %>% kable(digits = 3)
```


## Plot 3 year survival probability and median survival time 

To show the difference between estimating __ year survival and estimating the median survival time, I can plot these both on the same plot. 

```{r combined_plot}

# start with KM curve 
km1 + 
  # visualize 3 year survival (green)
  geom_segment(x = 3, xend = 3,
               y = 0, yend = three_yr_surv$surv[2], 
               size = 1, 
               arrow = arrow(),
               colour = "#5ec962") +
  geom_segment(x = 3, xend = 0,
               y = three_yr_surv$surv[2], yend = three_yr_surv$surv[2],
               size = 1, 
               arrow = arrow(length = unit(0.2, "inches")),
               colour = "#5ec962") + 
  # visualize median survival time (blue)
geom_segment(x = s1_median["median"], xend = s1_median["median"],
             # the first y-val must be slightly > 0 for the arrow to print correctly
             y = 0.08, yend = 0.5,
             size = 1,
             arrow = arrow(ends = "first"),
             colour = "#3b528b") + 
  geom_segment(x = 0, xend = s1_median["median"],
               y = 0.5, yend = 0.5,
               size = 1,
               arrow = arrow(ends = "last"),
               colour = "#3b528b")


```

Up to this point, we have been studying a Kaplan-Meier plot that describes the entire data set (all crown margin repairs). In practice, the objective is often to compare two subgroups from within the data set -- for instance, suppose we are interested in comparing how well crown margin repairs lasted between the root canal treated (RCT) and non-RCT groups. The plot below draws two Kaplan-Meier survival curves -- one for each of these subgroups. We notice that across time, the curve representing the RCT teeth is consistently below the curve representing the non-RCT teeth. This indicates that the curve for the RCT teeth is *dropping (decreasing) faster*, illustrating that the crown margin repairs done on RCT treated teeth do not last as long as the repairs done on non-RCT teeth. 

In addition to the curves in this graph, we also see the confidence intervals at each time point illustrated by the tinted area around each curve. The yellow and purple tinted areas overlap with each other quite a bit, which symbolizes that the difference between crown margin repairs done on RCT teeth and non-RCT teeth is subtle -- the repairs last only slightly longer on the non-RCT teeth.  

```{r compare_rct_groups}
km3 <- survfit2(Surv(time = Time, event = Event) ~ RCT, data = cmr) %>% 
  ggsurvfit() +
  labs(
    x = "Years (after repair)",
    y = "Overall probability that repair is intact"
  ) + 
  add_confidence_interval() + 
  add_risktable() + 
  scale_color_viridis(discrete = TRUE) + 
  scale_fill_viridis(discrete = TRUE) +
  theme_bw() 

km3
```


As a second example of comparing Kaplan-Meier plots between groups, let us suppose that we are working in a materials science context, where we are interested in comparing crown margin repairs that were done with glass ionomer (GI) to repairs done resin-modified glass ionomer (RMGI).  We see in the next figure that the survival curve representing the GI group is much lower than the curve for the RMGI group for all times after two years. We also see that the space between the two curves increases over time - the two curves are diverging. The confidence intervals do not overlap much at all after 2.5 years. These survival curves indicate that the crown margin repairs done with RMGI lasted notably longer than the repairs done with GI. There is evidence in this data set that the modification to GI makes a positive impact on the expected lifespan of crown margin repairs. 

```{r}
km4 <- survfit2(Surv(time = Time, event = Event) ~ Repair_Material,
                # NB: we are considering only those repairs that were done with GI and Amalgam
                data = cmr %>% filter(Repair_Material %in% c("RMGI", "GI"))) %>% 
  ggsurvfit() +
  labs(
    x = "Years (after repair)",
    y = "Overall probability that repair is intact"
  ) + 
  add_confidence_interval() + 
  add_risktable() + 
  scale_color_viridis(discrete = TRUE) + 
  scale_fill_viridis(discrete = TRUE) +
  theme_bw() 

km4
```

Our final survival analysis tool for this tutorial is a Cox proportional hazards model. This model examines each of the variables (i.e. the independent variables) in relationship the time-to-event outcome. A table summarizes the results of this Cox model using hazard ratios (HR), 95% confidence intervals (CI), and p-values. 


```{r cox_model}
coxph(Surv(time = Time, event = Event) ~ No_Cases + Age + Gender + CRA + Tooth_Type + Jaw +
        Repair_Material + Surfaces + No_Surfaces + RCT + Crown_Type + Clinic +
        Provider_Type,
      data = cmr) %>%
  tbl_regression(exp = TRUE)


```

## References 